---
title: "template_bayesian workflow"
format: html
editor: visual
---

# Purpose of This Document

This template guides you through a Bayesian analysis following McElreath's workflow from Statistical Rethinking. Each section includes explanatory notes that you should replace with your own content.

# \[Insert Model Name\]

## Load Packages

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(brms)
library(tidybayes)
library(bayesplot)
library(dagitty)
library(ggdag)
```

## Domain Knowledge

Previous Research Summarize key findings from previous research. What do we already know about this system?

Mechanisms Describe the potential mechanisms at work. How might your variables be causally related?

Expected Patterns Based on theory and previous research, what patterns do you expect to see in your data?

### Research Questions

State your research questions clearly and precisely. Good research questions should be:

Specific and well-defined Answerable with your available data

Connected to your theoretical framework :::

## Causal Model

```{r}

#| label: dag
#| fig-cap: "Directed Acyclic Graph representing hypothesized causal relationships"
#| code-fold: show

# Create your DAG
# Example:
dag <- dagitty('dag{
  X -> Y
  Z -> X
  Z -> Y
}')

# Plot the DAG
ggdag(dag) +
  theme_dag()
```

### DAG Interpretation

Explain what each node represents and why you've drawn the arrows as you have. What assumptions are embedded in this diagram?

## Data Simulation

### Simulate Data Implied by Causal Model

```{r}

```

## Data Loading and Processing

Data Loading and Processing

```{r}

# Load your data
# Example:
# data <- read_csv("your_data.csv")

# Document each cleaning step

```

## Model Building

### 

### Statistical Model Specification

$$
1 + 1
$$

### Computational Model Specification

BRMS or Stan model

```{r}

```

### Prior Predictive Checks

```{r}
#| label: prior-checks
#| fig-cap: "Prior predictive simulations"

# Simulate from priors
# Example:
n_sims <- 100
alpha_prior <- rnorm(n_sims, 0, 10)
beta_prior <- rnorm(n_sims, 0, 2)

# Plot simulations

```

### Model Fitting

```{r}

#| label: model-fit
#| cache: true

# Fit your model
# Example using brms:
# model <- brm(
#   y ~ x,
#   data = data,
#   family = gaussian(),
#   prior = c(
#     prior(normal(0, 10), class = Intercept),
#     prior(normal(0, 2), class = b),
#     prior(exponential(1), class = sigma)
#   ),
#   chains = 4,
#   cores = 4
# )

# Example using cmdstanr:
# model <- cmdstan_model("model.stan")
# fit <- model$sample(data = data, chains = 4, parallel_chains = 4)


```

## Model Checking

### Computational Diagnostics

```{r}

```

### Posterior Predictive Checks

```{r}


```

## Model Interpretation

### Parameter Estimates

```{r}

```

### Posterior Predictions

```{r}


```

### Predictor Residual Analysis

```{r}

```

### Sensitivity Analysis

```{r}

```

### Counterfactual Analysis

```{r}

```

## Conclusion

### Summary of Findings

### Key Results
Summarize your main findings here, connecting them back to your research questions.

### Limitations and Next Steps

Important Considerations
Discuss key limitations and potential future directions.

# References

